---
title: "Week 4 Notes: GLMs in Practice"
format:
  pdf:
    include-in-header:
      - file: macros.tex
    keep-tex: true
number-depth: 1
number-sections: true
---

This week, we'll mainly be doing some open-ended applications of generalized
linear models on real datasets. Some of them may involve doing some independent
work on your own in figuring out how to fit the models. For each question, I
would like you to consider the following issues:

- What modeling strategy is appropriate for the data?

For the purpose of downloading and cleaning the data, we will use the
`tidyverse` package:

```{r load, message = FALSE, warning = FALSE}
library(tidyverse)
```

# Problem 1: Estimating a Generalized Linear Model from Scratch

The following question is adapted from James Scott's SDS 383D course notes (all
mistakes my own).

The regression coefficients $\beta$ in a GLM are typically fit using some
variation on likelihood-based inference. In class, we saw that the score
function is of the form

$$
  s(\beta, \phi) = \sum_i \frac{\omega_i (Y_i - \mu_i)}{\phi V(\mu_i) \, g'(\mu_i)} X_i.
$$

which, under the canonical link, simplifies to 

$$
  s(\beta) = \sum_i \frac{\omega_i (Y_i - \mu_i)}{\phi} X_i.
$$

Let's take the specific case of logistic regression for a binomial outcome,
where $Z_i \sim \Binomial(N_i, \mu_i)$ for known sample size $N_i$ and $Y_i =
Z_i / N_i$ is the observed success fraction, where the canonical link is used.

A common approach to fitting GLMs via maximum likelihood (or, really, optimizing
*any* function) is to use *gradient descent*, i.e., iteratively update
\begin{align*}
  \beta \gets \beta + \gamma \, s(\beta)
\end{align*}
where $s(\cdot)$ is the score function and $\gamma$ is a "learning rate" that
is, ideally, selected automatically. Read up on the method of steepest
descent\thanks{if you want a textbook reference, see \emph{Numerical
Optimization} by Nocedal and Wright} , i.e. gradient descent and write your own
function that will fit a logistic regression by gradient descent. For extra
coding brownie points, try to maintain some level of generality to your code so
that it could also work with different GLMs, assuming you wrote different
sub-routines.

After doing this, grab the data `wdbc.csv` from the course website, or obtain
some other real data that interests you, and test out your routine (comparing it
to the output of, say, `glm` in `R` to sanity check your results). The WDBC file
has information on 569 breast-cancer patients from a study done in Wisconsin.
The first column is a patient ID, the second column is a classification of a
breast cell (Malignant or Benign), and the next 30 columns are measurements
computed from a digitized image of the cell nucleus. These are things like
radius, smoothness, etc. For this problem, use the first 10 features for X, i.e.
columns 3-12 of the file.

**Some notes:**

- We're trying to maximize the log-likelihood but the convention in the
  optimization literature is to minimize things. No big deal; what we're doing
  is the same as minimizing the negative of the log-likelihood.

- You need to add an intercept term, and the simplest way is to add a column of
  1’s as the first column of the feature matrix $\bX$. (If you've never seen
  this trick before, convince yourself why it makes sense.)

- Ensure that, at every iteration of gradient descent, you compute and store the
  current value of the log-likelihood so that you can track and plot the
  convergence of the algorithm.

- How did you select the learning rate $\gamma$? You can be as clever (or as
  non-clever) as you want here. Generally, smaller step sizes will be more
  robust, but converge slower, while larger step sizes may overshoot the
  solution entirely. One idea, called "line search" is to find the stepsize that
  itself minimizes the log-likelihood at each iteration, but this involves extra
  work (both in terms of coding and computations).

- How do you determine (automatically) that you have reached a reasonable
  stopping point? One possibility is to stop when $s(\beta)$ is "small enough"
  but you will need to be precise about what this means.

# Problem 2: Estimating a Generalized Linear Model from Scratch: 2nd Order Methods

Gradient descent is known as a _first order_ method for optimizing a function:
to implement it, we only need to evaluate the function itself and its partial
derivatives. A more powerful (but less generally applicable) approach is to use
a _second order_ method, which also makes use of the second derivatives, i.e.,
the Hessian matrix of the log-likelihood (or similar). Equivalently, the Hessian
is the *negative* of the observed Fisher information, which we derived in class.

a. Consider a point $b \in \Reals^P$ that serves as an intermediate guess for
   $\beta$. Show that, for any GLM, the second-order Taylor approximation of the
   log-likelihood around $b$ can be expressed in the form
   \begin{align*}
     q(\beta, b)
     = -\frac{1}{2}(\widehat{\bY} - \bX\beta)^\top W (\widehat{\bY} - \bX \beta)
         + c,
   \end{align*}
   where $W$ is a diagonal matrix of "working weights" and $\widehat{\bY}$ is a
   vector of "working responses", and $c$ is a constant that doesn't involve
   $\beta$. What is the matrix $W$ given by, and what are its diagonal entries?

a. Read up on Newton's method for optimizing smooth functions (e.g., in Nocedal
   and Wright, Chapter 2). In the above case, notice that Newton's method just
   corresponds to optimizing $q(\beta, \beta_t)$ with respect to $\beta$ in
   order to obtain the next value of $\beta_{t+1}$, and that this optimization
   is essentially a weighted least-squares problem. Implement it for the
   logistic regression model and test it out on the same data set you just used
   to test out gradient descent. Note: while you could do linear search, there
   is a "natural" step size of 1 in Newton's method. Verify that your numerical
   solution replicates the $\beta$ estimate you get when using a package solver,
   e.g., the `glm` function in `R`, up to minor numerical differences.

**Note:** This algorithm is referred to as *iteratively reweighted least
squares* (IRLS). A neat trick that makes IRLS very nice is that, in addition to
avoiding specifying a step size, there is also a very simple way to initialize
the algorithm: just set $\widehat{\bY} = \bY$.



# Problem 3: The Crabs Data

The following two datasets are from the textbook _Categorical Data Analysis_ by
Alan Agresti. The first dataset looks at the mating behavior of crabs during
spawning season. Roughly speaking, during spawning season some number of
"satellite" males cluster around females and try to fertilize their eggs. The
number of satellites a female attracts depends on a number of other traits of
the crab including her size, health, and color. The dataset can be loaded by
running the following code:

```{r crabs-load, cache = TRUE}
crabs_file <- str_c("https://raw.githubusercontent.com/alanagresti/",
                     "categorical-data/master/Crabs.dat")
crabs <- read_table(crabs_file) %>%
  mutate(spine = factor(spine), color = factor(color))
print(head(crabs))
```

The dataset contains the following columns:

- `crab`: an index for each crab (not really useful for us).
- `sat`: the number of satellites of each crab.
- `y`: 1 if the crab has a satellite, 0 otherwise.
- `width`: the width of the crab carapice in centimeters.
- `weight`: the weight in kilograms of the crab.
- `color`: A factor with levels 1 = light medium, 2 = medium, 3 = dark medium, 4
  = dark.
- `spine`: A factor with levels 1 = both good, 2 = one worn or broken, 3 = both
  worn or broken.

a. For each of the variables width, weight, color, and spine, make a prediction
   about how these features relate to the number of satellites within the
   population of all crabs.
   
b. Fit a Poisson loglinear model (could be as simple or complex as you want in
   terms of, e.g., interactions, but be prepared to defend any choices you
   make). Then, interpret the estimated coefficients in terms of what they say
   about the relationship between each predictor and the number of satellites.
   
c. Use the loglinear model to assess each of the predictions you made, both in
   terms of whether the data supports (or fails to support) the predictions and
   the strength of the evidence in terms of uncertainty quantification.

d. A possible limitation of the Poisson loglinear model is that count data is
   often *overdispersed* with respect to the Poisson distribution, i.e., we tend
   to see $\Var(Y_i \mid X_i = x) > \E(Y_i \mid X_i = x)$. The main consequence
   of overdispersion is that the model will tend to give a false sense of
   precision in the inferences/predictions it makes.
   
   Try to come up with a simple procedure for assessing whether or not the
   loglinear model you fit is overdispersed relative to the Poisson
   distribution. Then, use this to determine if there is evidence of
   overdispersion for the crab dataset. _Hint:_ there are a lot of possibilities
   here, but one thing to observe is that if the Poisson model is correct that
   we should have $\E\{(Y_i - \mu_i)^ 2 / \mu_i\} = 1$.

# Problem 4: Injury Dataset

From Agresti this dataset contains 

> ... results of accidents in the state of Maine for 68,694 passengers in autos
> and light trucks. The table classifies passengers by gender, location of
> accident, seat-belt use, and injury.

The columns `yes` and `no` state whether the passenger was injured in the
accident.

```{r injury-load}
injury_file <- str_c("https://raw.githubusercontent.com/alanagresti/",
                     "categorical-data/master/Injury_binom.dat")
injury <- read_table(injury_file)[,-6]
head(injury)
```

In cases like this, where the data has columns with counts of "yes" and "no"
rather than columns of 0s and 1s, a standard way to fit a logistic regression
model is to have a response be a matrix with the success counts in the first
column and failure counts in the second. For example:

```{r seatbelt-glm}
seatbelt_glm <- glm(cbind(yes, no) ~ seatbelt * gender,
                    data = injury,
                    family = binomial)
```

fits a logistic regression model the models $\logit p = \beta_0 + \beta_1 \times
\texttt{seatbelt\_yes} + \beta_2 \times \texttt{gender\_male} + \beta_3 \times
\texttt{seatbelt\_yes} \times \texttt{gender\_male}$.

Use a logistic regression model(s) to attempt to answer the following questions:

a. Within the subpopulations defined by each (gender, location) combination, is
   there evidence that use of a seatbelt results in reduced incidents of injury?
   If so, can you give an estimate or this effect and an estimate about how
   precisely the effect is measured?

a. Is there evidence that, within each subpopulation defiend by an individual's
   location, the effect of using a seatbelt is different for males and females?
   If so, can you give an estimate of this effect (e.g., seatbelt use reduces
   the odds of an injury for women by a factor of XXX more/less relative to men)
   and an estimate of how precisely this is measured?

